|

### 2022-02-08 submission/manuscript.rmd
almost final version after comments

---
title: |
 Individualized treatment effect was predicted best by modeling 
 baseline risk in interaction with treatment assignment
abstract: |
 **Objective**: To compare different risk-based methods for optimal
 prediction of individualized treatment effects from RCTs. **Study
 Design and Setting**: We simulated RCT data using diverse assumptions
 for the average treatment effect, a baseline prognostic index of risk
 (PI), the shape of its interaction with treatment (none, linear,
 quadratic or non-monotonic) and the magnitude of treatment-related
 harms (none or constant independent of the PI). In each sample we
 predicted absolute benefit using: models with a constant relative
 treatment effect; stratification in quarters of the PI; models
 including a linear interaction of treatment with the PI; models
 including an interaction of treatment with a restricted cubic spline
 (RCS) transformation of the PI; an adaptive approach using Akaike’s
 Information Criterion. We evaluated predictive performance using root
 mean squared error and measures of discrimination and calibration for
 benefit. **Results**: The linear-interaction model and the
 RCS-interaction model outperformed the constant treatment effect
 model in many simulation scenarios. The RCS-model was optimal when
 quadratic or non-monotonic deviations from a constant treatment
 effect were stronger, and when sample size was larger. Larger sample
 size also supported the adaptive approach. Illustrations in the
 GUSTO-I trial confirmed these findings. **Conclusion**: An
 interaction between baseline risk and treatment assignment should be
 considered to improve treatment effect predictions. Non-linear RCS
 interactions should be considered only in larger sample sizes.
author:
  - name: Alexandros Rekkas
    affiliation: 1
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: David M. Kent
    affiliation: 2
  - name: Ewout W. Steyerberg
    affiliation: 3
  - name: David van Klaveren
    affiliation: 4
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
  - code: 3
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 4
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::elsevier_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

A *stratified HTE method* has been suggested as an alternative to traditional
subgroup analyses. Patients are stratified into equally-sized risk strata---in
this case based on risk quartiles. Absolute treatment effects within risk strata
are estimated by the difference in event rate between patients in the control
arm and patients in the treated arm. We considered this approach as a reference,
expecting it to perform worse than the other candidates, as its objective is to
provide an illustration of HTE rather than to optimize individualized benefit
predictions.

@@

@@

We demonstrate the different methods for individualizing treatment benefits
using data from `r nrow(gusto)` patients with acute myocardial infarction (MI)
included in the GUSTO-I trial.
`r treatmentArms %>% filter(treatment == 1) %>% select(n)` 
patients were randomized to tissue plasminogen activator (tPA) treatment and
`r treatmentArms %>% filter(treatment == 0) %>% select(n)`
were randomized to streptokinase. The outcome of interest was 30-day mortality
(total of `r sum(gusto$outcome)` events), recorded for all patients.

@@

The linear interaction model outperformed all RCS methods in terms of RMSE in
scenarios with true constant relative treatment effect (OR = 0.8, N = `r 4250`
and AUC = 0.75), strong linear and even strong quadratic deviations from a
constant relative treatment effect (Figure \ref{fig:rmsebase}; panels
A-C). However, with non-monotonic deviations from a constant relative treatment
effect, the RMSE of the linear interaction model increased substantially,
especially in the presence of treatment-related harms (Figure
\ref{fig:rmsebase}; panel D). In these scenarios, RCS-3 outperformed all other
methods in terms of RMSE. As might be expected the constant treatment effect
approach had overall best performance under true constant treatment effect
settings. It was sensitive to all considered deviations, resulting in increased
RMSE. Finally, the adaptive approach had comparable performance to the
best-performing method in each scenario. However, in comparison with the
best-performing approach, its RMSE was more variable in the scenarios with
linear and non-monotonic deviations, especially when also including moderate or
strong treatment-related harms. On closer inspection, we found that this
behavior was caused by wrongly selecting the constant
treatment effect model in a substantial proportion of the replications
(Supplement, Figure S1). This problematic behavior was less with larger sample
sizes (see below).

@@

Increasing the sample size to `r 17000` favored RCS-3 the most, It achieved
lowest or close to lowest RMSE across all scenarios (Figure
\ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
non-monotonic deviations RCS-3 had lower RMSE (median
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for strong quadratic deviations and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for non-monotonic deviations with no treatment-related harms) compared
to the linear interaction approach (median 
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)` and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`,
respectively), regardless of the strength of treatment-related
harms. Due to the large sample size, the RMSE of the adaptive approach
was even more similar to the best-performing method, and the constant
relative treatment effect model was less often wrongly selected
(Supplement, Figure S2).

```{r rmsesamplesize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in simulated samples of size 17,000 rather than 4,250 in Figure \\ref{fig:rmsebase}", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
```

When we increased the AUC of the true prediction model to 0.85 (OR = 0.8 and N =
`r 4250`). RCS-3 had the lowest RMSE in the case of strong quadratic or
non-monotonic deviations and very comparable performance to the -- optimal --
linear interaction model in the case of strong linear deviations (median RMSE
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)` for RCS-3 compared to 
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`
for the linear interaction model). As observed in the base case scenario the
adaptive approach wrongly selected the constant treatment effect model
(`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
and
`r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
of the replications in the strong linear and non-monotonic deviation scenarios
without treatment-related harms, respectively), leading to more variability of
the RMSE (Supplement, Figure S3).

```{r rmseauc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated samples 4,250. True prediction AUC of 0.85.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_auc.tiff")))
# knitr::include_graphics(here::here("figures/rmse_auc.png"))
```

In comparison with the true approach, dsicrimination for benefit in the scenario
with a constant relative treatment effect was only slightly lower for the linear
interaction model, but substantially lower for the non-linear RCS approaches
(Figure \ref{fig:discrimination}; panel A). With strong linear or quadratic
deviations from a constant relative treatment effect, all methods discriminated
quite similarly (Figure \ref{fig:discrimination}; panels B-C). In the scenario
with non-monotonic deviations, the constant effect model had much lower
discriminative ability compared to all other methods (median AUC of
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
for the constant effects model,
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_linear)`
for the linear interaction model and
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`
for the best-performing RCS-3; Figure \ref{fig:discrimination}; panel D). The
adaptive approach was unstable in terms of discrimination for benefit,
especially in the presence of treatment-related harms. With increasing number of
RCS knots, we observed decreasing median values and increasing variability of
the c-for-benefit in all scenarios. When we increased the sample size to 17,000
we observed similar trends, however the performance of all methods was more
stable (Supplement, Figure S4). Finally, when we increased the true prediction
AUC to 0.85 the adaptive approach in the case of non-monotonic deviations was,

again, quite unstable, especially with null or moderate treatment-related harms
(Supplement, Figure S5). In these scenarios, the adaptive approach tended to
select more often the inferior constant treatment effect method (Supplement,
Figure S3)

@@

In terms of calibration for benefit, the constant effects model outperformed all
other models in the scenario with true constant treatment effects, but was
miscalibrated for all deviation scenarios (Figure \ref{fig:calibration}). The
linear interaction model showed best or close to best calibration across all
scenarios and only showed worse calibration compared to RCS-3 in case of
non-monotonic deviations and treatment-related harms (Figure
\ref{fig:calibration}; panel D). The adaptive approach was worse calibrated in
scenarios with strong linear and non-monotonic deviations compared to the linear
interaction model and RCS-3. When we increased sample size to 17,000 similar
conclusions on calibration for benefit could be drawn. As expected, all methods
displayed more stable calibration performance due to the larger number of
patients (Supplement, Figure S6). When we increased the true prediction AUC to
0.85, the linear interaction model was worse calibrated, on average, than RCS-3
in the case of strong quadratic deviations from constant relative treatment
effects (Supplement, Figure S7).

The results from all individual scenarios can be explored online at
[https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/).

@@

In line with previous analyses [@Califf1997; @Steyerberg2000], we fitted a
logistic regression model with 6 baseline covariates, i.e. age, Killip class,
systolic blood pressure, heart rate, an indicator of previous MI, and the
location of MI, to predict 30-day mortality risk. A constant effect of treatment
was included in the model. When deriving risk predictions for individuals we set
the treatment indicator to 0. More information on model development can be found
in the supplement (Section XX).

@@

We used the risk linear predictor to fit a constant treatment effect model, a
linear interaction model and a RCS-3 model individualizing absolute benefit
predictions. RCS-4 and RCS-5 models were excluded. In our simulations these
methods were always outperformed by the simpler approaches and were often
overfitted. Finally, an adaptive approach with only the 3 candidate models was
also applied.

@@

```{r gusto, cache=TRUE, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots. Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/gusto.tiff")))
# knitr::include_graphics(here::here("figures/gusto.png"))
```

@@

The adaptive approach had adequate performance, following closely on average the
performance of the “true” model in most scenarios. With smaller sample sizes it
tended to miss the treatment-risk interactions and selected simpler models
(Supplementary Table S7). This conservative behavior resulted in increased RMSE
variability in these scenarios, especially in the case of true strong linear or
non-monotonic deviations from the base case scenario. Therefore, in the case of
smaller sample sizes the simpler linear interaction model may be a safer choice
for predicting absolute benefits, if a non-constant treatment effect is
suspected.

Risk-based approaches to predictive HTE estimate treatment benefit as a function
of baseline risk. A limitation of our study is that we assumed treatment benefit
to be a function of baseline risk in the majority of the simulation
scenarios. We also considered constant moderate and strong treatment-related
harms, applied on the absolute scale to expand the range of scenarios in line
with previous work [@Glasziou1995]. In a limited set of scenarios where we
assumed the existence of true treatment-covariate interactions, our conclusions
remained unchanged. Even though the average error rates increased for all the
considered methods, due to the miss-specification of the outcome model, the
linear interaction model had the lowest error rates. RCS-3 had very comparable
performance. The constant treatment effect model often gave biased results,
espcially in the presence of moderate or strong treatment-related harms. All the
results of these simulations can be found in Supplement, Section XX. Future
simulation studies could explore the effect of more extensive deviations from
risk-based treatment effects.

>>>>

---
title: |
 Individualized treatment effect was predicted best by modeling 
 baseline risk in interaction with treatment assignment
abstract: |
 **Objective**: To compare different risk-based methods for optimal
 prediction of individualized treatment effects from RCTs. **Study
 Design and Setting**: We simulated RCT data using diverse assumptions
 for the average treatment effect, a baseline prognostic index of risk
 (PI), the shape of its interaction with treatment (none, linear,
 quadratic or non-monotonic) and the magnitude of treatment-related
 harms (none or constant independent of the PI). In each sample we
 predicted absolute benefit using: models with a constant relative
 treatment effect; stratification in quarters of the PI; models
 including a linear interaction of treatment with the PI; models
 including an interaction of treatment with a restricted cubic spline
 (RCS) transformation of the PI; an adaptive approach using Akaike’s
 Information Criterion. We evaluated predictive performance using root
 mean squared error and measures of discrimination and calibration for
 benefit. **Results**: The linear-interaction model and the
 RCS-interaction model outperformed the constant treatment effect
 model in many simulation scenarios. The RCS-model was optimal when
 quadratic or non-monotonic deviations from a constant treatment
 effect were stronger, and when sample size was larger. Larger sample
 size also supported the adaptive approach. Illustrations in the
 GUSTO-I trial confirmed these findings. **Conclusion**: An
 interaction between baseline risk and treatment assignment should be
 considered to improve treatment effect predictions. Non-linear 
 interactions should be considered only in larger sample sizes.
author:
  - name: Alexandros Rekkas
    affiliation: 1
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: David M. Kent
    affiliation: 2
  - name: Ewout W. Steyerberg
    affiliation: 3
  - name: David van Klaveren
    affiliation: 4
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
  - code: 3
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 4
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::elsevier_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

A *stratified HTE method* has been suggested as an alternative to traditional
subgroup analyses [@Kent2019; @PathEnE]. Patients are stratified into
equally-sized risk strata---in this case based on risk quartiles. Absolute
treatment effects within risk strata are estimated by the difference in event
rate between patients in the control arm and patients in the treated arm. We
considered this approach as a reference, expecting it to perform worse than the
other candidates, as its objective is to provide an illustration of HTE rather
than to optimize individualized benefit predictions.

@@

For each scenario setting we performed 500 replications, within which all the
considered models were fitted. For comparing between models we simulated a
super-population of size 500,000 for each scenario. We calculated RMSE and
discrimination and calibration for benefit of the models derived in each
replication of our simulation settings within this super-population. 
## Empirical illustration

@@

We demonstrated the different methods for individualizing treatment benefits
using data from `r nrow(gusto)` patients with acute myocardial infarction (MI)
included in the GUSTO-I trial.
`r treatmentArms %>% filter(treatment == 1) %>% select(n)` 
patients were randomized to tissue plasminogen activator (tPA) treatment and
`r treatmentArms %>% filter(treatment == 0) %>% select(n)`
were randomized to streptokinase. The outcome of interest was 30-day mortality
(total of `r sum(gusto$outcome)` events), recorded for all patients.

In line with previous analyses [@Califf1997; @Steyerberg2000], we fitted a
logistic regression model with 6 baseline covariates, i.e. age, Killip class,
systolic blood pressure, heart rate, an indicator of previous MI, and the
location of MI, to predict 30-day mortality risk. A constant effect of treatment
was included in the model. When deriving risk predictions for individuals we set
the treatment indicator to 0. More information on model development can be found
in the supplement (Section 6).

@@

The linear interaction model outperformed all RCS methods in terms of RMSE in
scenarios with true constant relative treatment effect (OR = 0.8, N = `r 4250`
and AUC = 0.75), strong linear and even strong quadratic deviations from a
constant relative treatment effect (Figure \ref{fig:rmsebase}; panels
A-C). However, with non-monotonic deviations from a constant relative treatment
effect, the RMSE of the linear interaction model increased substantially,
especially in the presence of treatment-related harms (Figure
\ref{fig:rmsebase}; panel D). In these scenarios, RCS-3 outperformed all other
methods in terms of RMSE. As might be expected the constant treatment effect
approach had overall best performance under true constant treatment effect
settings. It was sensitive to all considered deviations, resulting in increased
RMSE. Finally, the adaptive approach had comparable performance to the
best-performing method in each scenario. However, in comparison with the
best-performing approach, its RMSE was more variable in the scenarios with
linear and non-monotonic deviations, especially when also including moderate or
strong treatment-related harms. On closer inspection, we found that this
behavior was caused by wrongly selecting the constant
treatment effect model in a substantial proportion of the replications
(Supplement, Figure S3). This problematic behavior was less with larger sample
sizes (see below).

@@

Increasing the sample size to `r 17000` favored RCS-3 the most, It achieved
lowest or close to lowest RMSE across all scenarios (Figure
\ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
non-monotonic deviations RCS-3 had lower RMSE (median
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for strong quadratic deviations and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for non-monotonic deviations with no treatment-related harms) compared
to the linear interaction approach (median 
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)` and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`,
respectively), regardless of the strength of treatment-related
harms. Due to the large sample size, the RMSE of the adaptive approach
was even more similar to the best-performing method, and the constant
relative treatment effect model was less often wrongly selected
(Supplement, Figure S4).

```{r rmsesamplesize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in simulated samples of size 17,000 rather than 4,250 in Figure \\ref{fig:rmsebase}. RMSE was calculated on a super-population of size 500,000", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
```

When we increased the AUC of the true prediction model to 0.85 (OR = 0.8 and N =
`r 4250`). RCS-3 had the lowest RMSE in the case of strong quadratic or
non-monotonic deviations and very comparable performance to the -- optimal --
linear interaction model in the case of strong linear deviations (median RMSE
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)` for RCS-3 compared to 
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`
for the linear interaction model). As observed in the base case scenario the
adaptive approach wrongly selected the constant treatment effect model
(`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
and
`r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
of the replications in the strong linear and non-monotonic deviation scenarios
without treatment-related harms, respectively), leading to more variability of
the RMSE (Supplement, Figure S5).

```{r rmseauc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in simulated samples 4,250. True prediction AUC of 0.85. RMSE was calculated on a super-population of size 500,000", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_auc.tiff")))
# knitr::include_graphics(here::here("figures/rmse_auc.png"))
```

When assuming a true constant relative treatment effect, discrimination for
benefit was only slightly lower for the linear interaction model, but
substantially lower for the non-linear RCS approaches (Figure
\ref{fig:discrimination}; panel A). With strong linear or quadratic deviations
from a constant relative treatment effect, all methods discriminated quite
similarly (Figure \ref{fig:discrimination}; panels B-C). In the scenario with
non-monotonic deviations, the constant effect model had much lower
discriminative ability compared to all other methods (median AUC of
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
for the constant effects model,
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_linear)`
for the linear interaction model and
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`
for the best-performing RCS-3; Figure \ref{fig:discrimination}; panel D). The
adaptive approach was unstable in terms of discrimination for benefit,
especially in the presence of treatment-related harms. With increasing number of
RCS knots, we observed decreasing median values and increasing variability of
the c-for-benefit in all scenarios. When we increased the sample size to 17,000
we observed similar trends, however the performance of all methods was more
stable (Supplement, Figure S6). Finally, when we increased the true prediction
AUC to 0.85 the adaptive approach in the case of non-monotonic deviations was,
again, more conservative, especially with null or moderate treatment-related
harms (Supplement, Figure S7).

@@

In terms of calibration for benefit, the constant effects model outperformed all
other models in the scenario with true constant treatment effects, but was
miscalibrated for all deviation scenarios (Figure \ref{fig:calibration}). The
linear interaction model showed best or close to best calibration across all
scenarios and only showed worse calibration compared to RCS-3 in case of
non-monotonic deviations and treatment-related harms (Figure
\ref{fig:calibration}; panel D). The adaptive approach was worse calibrated in
scenarios with strong linear and non-monotonic deviations compared to the linear
interaction model and RCS-3. When we increased sample size to 17,000 similar
conclusions on calibration for benefit could be drawn. As expected, all methods
displayed more stable calibration performance due to the larger number of
patients (Supplement, Figure S8). When we increased the true prediction AUC to
0.85, the linear interaction model was worse calibrated, on average, than RCS-3
in the case of strong quadratic deviations from constant relative treatment
effects (Supplement, Figure S9).

The results from all individual scenarios can be explored online at
[https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/). Additionally,
all the code for the simulations can be found at
[https://github.com/rekkasa/arekkas_HteSimulation_XXXX_2021](https://github.com/rekkasa/arekkas_HteSimulation_XXXX_2021)

@@

@@

We used the derived prognostic index to fit a constant treatment effect model, a
linear interaction model and a RCS-3 model individualizing absolute benefit
predictions. RCS-4 and RCS-5 models were excluded. In our simulations these
methods were always outperformed by the simpler approaches and were often
overfitted. Finally, an adaptive approach with only the 3 candidate models was
also applied.

@@

```{r gusto, cache=TRUE, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3 knots. Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/gusto.tiff")))
# knitr::include_graphics(here::here("figures/gusto.png"))
```

@@

The adaptive approach had adequate performance, following closely on average the
performance of the “true” model in most scenarios. With smaller sample sizes it
tended to miss the treatment-risk interactions and selected simpler models
(Supplement Section 4). This conservative behavior resulted in increased RMSE
variability in these scenarios, especially in the case of true strong linear or
non-monotonic deviations from the base case scenario. Therefore, in the case of
smaller sample sizes the simpler linear interaction model may be a safer choice
for predicting absolute benefits, if a non-constant treatment effect is
suspected.

Risk-based approaches to predictive HTE estimate treatment benefit as a function
of baseline risk. A limitation of our study is that we assumed treatment benefit
to be a function of baseline risk in the majority of the simulation
scenarios. We also considered constant moderate and strong treatment-related
harms, applied on the absolute scale to expand the range of scenarios in line
with previous work [@Glasziou1995]. In a limited set of scenarios where we
assumed the existence of true treatment-covariate interactions, our conclusions
remained unchanged. Even though the average error rates increased for all the
considered methods, due to the miss-specification of the outcome model, the
linear interaction model had the lowest error rates. RCS-3 had very comparable
performance. The constant treatment effect model often gave biased results,
espcially in the presence of moderate or strong treatment-related harms. All the
results of these simulations can be found in Supplement, Section 6. Future
simulation studies could explore the effect of more extensive deviations from
risk-based treatment effects.
